\sys: Controllable Program Regeneration

### 2025-06-10 paper/draft.tex
Refine intro
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

Modern software systems are in constant flux.
Developers refactor legacy code,
adapt libraries to evolving APIs, translate modules across programming
languages, and harden components against emerging security threats.
Each transformation demands a careful balance between preserving intended
functionality and satisfying new goals such as modularity, performance, or
safety.
While these transformations are conceptually simple, they are often
brittle, time-consuming, and require deep expertise.
% TODO: Introduce LLMs around here

@@

This paper present \sys, a system for scalable program regeneration.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties---such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure. Given a source program and
such a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

The paper's contributions are:
\begin{itemize}
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program;
 \item A property-aware transformation framework, combining logic-based reasoning, minimal extraction, and LLM-backed synthesis;
 % TODO: Implementation of sys?
\end{itemize}

By coupling logic-driven reasoning with modern synthesis tools, \sys enables a new class of transformation workflows---ones that are expressive, verifiable, and adaptable across software tasks.

>>>>

Modern software systems are in constant flux.
Developers refactor legacy code~\cite{Fowler99,Mens04},
adapt libraries to evolving APIs,
translate modules across programming languages,
and harden components against emerging security threats.
These transformations are brittle, time-consuming, and may require deep expertise,
even if the can be conceptually simple.
% TODO: Introduce LLMs around here

Recent advances in large language models (LLMs) have shown great promise in
program synthesis and transformation tasks.
Yet LLMs suffer from several well-documented limitations:
they struggle with negative reasoning or the proof of absence properties, their outputs often over-align with input patterns—making
them vulnerable to adversarial or malformed prompts, and it
remains difficult to enforce rich semantic or domain-specific constraints on
generated code.
As a result, while LLM-backed tools can
accelerate prototyping, developers still lack reliable ways to verify or steer
their transformations to meet precise correctness or security requirements.

@@

This paper present \sys, a system for \emph{scalable program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties---such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure. Given a source program and
such a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

The paper's contributions are:
\begin{itemize}
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program;
 \item A property-aware transformation framework, combining logic-based reasoning, and LLM-backed synthesis;
 \item An empirical evaluation demonstrating \sys’s effectiveness and scalability across transformation scenarios.
\end{itemize}

By coupling logic-driven reasoning with modern synthesis tools, \sys enables a
new class of transformation workflows---ones that are expressive, to some extent verifiable,
and adaptable across software tasks.
